Summary of the things that I learned

### 2024-09-27 200-category-theory.tex
Make notation in category-theory more consistent

Now, in $ \SET $, for a family $ (Y_x)_x $, consider the dependent product $ \prod_{x : X} Y_x $. Its elements $ (y_x)_x $ can be identified with morphisms from the terminal set: $ \{ \star \} \to \prod_{x : X} Y_x $, sending $ \star $ to $ (y_x)_x $. However, they can also be identified with the morphisms $ f: X \to \sum_{x : X} Y_x $ that make the following diagram commute, sending $ x $ to $ y_x $:
\begin{center}
  \begin{tikzcd}
    X \arrow[rd, "\id X"'] \arrow[rr, "f"] & & \sum_{x : X} Y_x \arrow[ld, "\pi_1"]\\
    & X
  \end{tikzcd}
\end{center}
These are morphisms in $ (\SET \downarrow X) $ from $ (X, \id X) $ to $ (\sum_{x : X} Y_x, \pi_1) $. Note that for the terminal morphism $ f: X \to \{ \star \} $, we have $ (X, \id X) = f^*(\{\star\}, \id{\{\star\}}) $. To summarize, we have an equivalence
\[ (\SET \downarrow X)\left(f^*(\{ \star \}, \id{\{\star\}}), \left(\sum_{x : X} Y_x, \pi_1 \right) \right) \simeq (\SET \downarrow {\{\star\}})\left( (\{\star\}, \id{\{ \star \}}), \left( \prod_{x : X} Y_x, ! \right) \right) \]
for $ ! $ the terminal morphism. Now, for an internal indexed family $ f : Y \to X $ and given a family of families $ ((Z_y)_{y : Y_x})_{x : X} $, we can wonder whether we can construct the family of dependent products $ (\prod_{y : Y_x} Z_y)_x $. In $ \SET $, this is definitely possible, and from this, we get an equivalence again
\[ \left(\SET \downarrow \sum_{x : X} Y_x \right)\left(f^*(X, \id X), (\sum_{x : X} \sum_{y : Y_x} Z_y, p) \right) \simeq (\SET \downarrow A)\left((X, \id X), \left(\sum_{x : X} \prod_{y : Y_x} Z_y, \pi_1\right)\right). \]
with $ p : \sum_{x : X} \sum_{y : Y_x} Z_y \to \sum_{x : X} Y_x $ defined as $ p(x, y, z) = (x, y) $. These equivalences suggest an adjunction $ f^* \dashv \prod_{\dots} $. We can use this to define in general
\begin{definition}
  For a category $ C $ and a morphism $ f: Y \to X $, the \iindex{dependent product} along $ f $ is, if it exists, the right adjoint to the pullback functor:
  \begin{center}
    \begin{tikzcd}
      (C \downarrow A) \arrow[r, shift left=2, "f^*"{name=A}] &
      (C \downarrow B) \arrow[l, shift left=2, "\prod_f"{name=B}]
      \ar[from=A, to=B, symbol=\dashv]
    \end{tikzcd}
  \end{center}
\end{definition}
\begin{remark}
  As argued above, we can recover the familiar dependent product $ \prod_{x : X} Y_x $ of a family $ (Y_x)_x $ as the dependent product $ \prod_f (\sum_x Y_x, \pi_1) $ along the terminal morphism $ f: X \to I $. Here we use the equivalence between $ (C \downarrow I) $ and $ C $.
\end{remark}

Now we turn our attention to dependent sums. In $ \SET $, let $ (X_a)_a $ and $ (B_a)_a $ be two families over $ A $ and let $ ((Y_b)_{b : B_a})_{a : A} $ be a family of families. Let $ \alpha: \sum_{a : A} B_a \to A $ be the internal representation of $ (B_a)_a $. A family of maps $ f_a : (\sum_{b : B_a} Y_b)_a \to X_a $ consists of maps $ Y_b \to X_a $ for all $ b : B_a $, so these are maps $ f_b : Y_b \to X_{\alpha(b)} $. This gives an equivalence
\[ (\SET \downarrow A)((\sum_{b : B_a} Y_b)_a, (X_a)_a) \simeq (\SET \downarrow (\sum_{a : A} B_a))((Y_b)_b, \alpha^*((X_a)_a)). \]
This, again, suggests an adjunction which we will use as a definition.
\begin{definition}
  For a category $ C $ and a morphism $ \alpha: B \to A $, the \iindex{dependent sum} along $ \alpha $ is, if it exists, the left adjoint to the pullback functor:
  \begin{center}
    \begin{tikzcd}
      (C \downarrow A) \arrow[r, shift right=2, "\alpha^*"'{name=B}] &
      (C \downarrow B) \arrow[l, shift right=2, "\sum_\alpha"'{name=A}]
      \ar[from=A, to=B, symbol=\dashv]
    \end{tikzcd}
  \end{center}
\end{definition}

However, note that the conversion from an external to an internal representation in $ \SET $ already contained a dependent sum, which is no coincidence. It turns out that in practice, we will never have a hard time obtaining dependent sums:
\begin{lemma}\label{lem:sum-postcomposition}
  Let $ \alpha : C(B, A) $ be a morphism in a category. If the pullback functor $ \alpha^*: (C \downarrow A) \to (C \downarrow B) $ exists, it has a left adjoint given by postcomposition with $ \alpha $.
\end{lemma}
\begin{proof}
  For morphisms $ f: X \to A $, $ g: Y \to B $, the universal property of the pullback, with the following diagram
  \begin{center}
    \begin{tikzcd}
      Y \arrow[r, dashed, "\varphi"'] \arrow[rr, "\psi", bend left] \arrow[dr, "g"'] & \alpha^* X \arrow[d, "\alpha^* f"'] \arrow[r] \arrow[rd, phantom, "\lrcorner", very near start] & X \arrow[d, "f"]\\
      & B \arrow[r, "\alpha"'] & A
    \end{tikzcd}
  \end{center}
  gives an equivalence between morphisms $ \varphi: Y \to \alpha^* X $ that commute with $ g $ and $ \alpha^* f $, and morphisms $ \psi: Y \to X $ that commute with $ g $, $ f $ and $ \alpha $. In other words:
  \[ (C \downarrow A)(g \cdot \alpha, f) \simeq (C \downarrow B)(g, \alpha^*(f)), \]
  which shows the adjunction.
\end{proof}

Now, let $ f : X \to A $ be the internal representation of an indexed family $ (X_a)_a $ and let $ \alpha : A \to I $ be the terminal projection. We have $ \sum_{a : A} X_a = f \circ \alpha : X \to I $. By the equivalence between $ (C \downarrow I) $ and $ C $, we see that the dependent sum of the family $ (X_a)_a $ is exactly $ X $. Therefore, our attention is mainly focused on the dependent product.

@@

\begin{definition}
  A \index{cartesian closed!locally}\textit{locally cartesian closed} category is a category $ C $ with pullbacks such that each pullback functor $ \alpha^* $ has a right adjoint.
\end{definition}

Apart from having dependent sums and products, there also is the following theorem that shows the significance of locally cartesian closedness:
\begin{lemma}\label{lem:locally-cartesian-closed}
  A category $ C $ is locally cartesian closed iff $ (C \downarrow A) $ is cartesian closed for each $ A : C $.
\end{lemma}
\begin{proof}
  See the end of Section 1.3 of \autocite{freyd}.
\end{proof}

\begin{remark}\label{rem:pullback-of-projection}
  Note that for $ X, Y, Z : C $ and $ f: C(Y, X) $, the following diagram shows that $ f^* \Delta_X Z \cong \Delta_Y Z $:
  \begin{center}
    \begin{tikzcd}
      Y \times Z \arrow[d, "p_1"] \arrow[r, "f \times \id Z"] & X \times Z \arrow[d, "p_1"] \arrow[r, "p_2"] & Z \arrow[d, "!"]\\
      Y \arrow[r, "f"] & X \arrow [r, "!"] & T
    \end{tikzcd}
  \end{center}
\end{remark}

\begin{lemma}\label{lem:constant-dependent-product}
  For $ W, X, Z : C $ and $ p_1 : X \times Z \to X $,
  \[ \prod_{p_1} \Delta_{X \times Z} W \cong \Delta_X W^Z \]
\end{lemma}
\begin{proof}
  First of all, note that $ (C \downarrow X \times Z) \cong ((C \downarrow X) \downarrow \Delta_X Z) $. Also note that the composite morphism $ (X \times Z) \times W \xrightarrow{p_1} X \times Z \xrightarrow{p_1} X $ is the element $ \Delta_X (W \times Z) : (C \downarrow X) $.

  By Proposition 1.34 in \autocite{freyd}, $ \prod_{p_1} \Delta_{X \times Z} W $ is given as the following pullback:
  \begin{center}
    \begin{tikzcd}
      \prod_{p_1} \Delta_{X \times Z} W \arrow[r] \arrow[d] & (\Delta_X Z \times W)^{\Delta_X Z} \arrow[d, "(\Delta_X p_1)^{\Delta_X Z}"]\\
      X \arrow[r] & (\Delta_X Z)^{\Delta_X Z}
    \end{tikzcd}
  \end{center}

  By Lemma \ref{lem:delta-exponentials}, $ \Delta_X $ preserves exponential objects, so the morphism on the right is $ \Delta_X p_1^Z : (C \downarrow X)(\Delta_X (Z \times W)^Z, \Delta_X Z^Z) $. However, we have an isomorphism $ (Z \times W)^Z \cong Z^Z \times Z^W $, and then the morphism on the right becomes
  \[ \Delta_X p_1 : (C \downarrow X)(\Delta_X(Z^Z \times Z^W), \Delta_X Z^Z) \]
  We also have an isomorphism $ X \cong \Delta_X T $. Then by Lemma \ref{lem:delta-limits} and Remark \ref{rem:pullback-of-projection}, the pullback of this diagram is $ \Delta_X W^Z $:
  \begin{center}
    \begin{tikzcd}
      \Delta_X W^Z \arrow[r] \arrow[d] & \Delta_X (Z^Z \times W^Z) \arrow[d, "\Delta_X p_1"]\\
      \Delta_X T \arrow[r] & \Delta_X Z^Z
    \end{tikzcd}
  \end{center}
\end{proof}

\section{(Weakly) terminal objects}
\begin{definition}
  If a category has an object $ t $, such that there is a (not necessarily unique) morphism to it from every other object in the category, $ t $ is said to be a \iindex{weakly terminal object}.
\end{definition}

\begin{definition}
  Let $ C $ be a category with terminal object $ t $. For an object $ c: C $, a \iindex{global element} of $ c $ is a morphism $ f: C(t, c) $.
\end{definition}

>>>>

Now, in $ \SET $, for a family $ (Y_x)_x $, consider the dependent product $ \prod_{x : X} Y_x $. Its elements $ (y_x)_x $ can be identified with morphisms from the terminal set: $ \{ \star \} \to \prod_{x : X} Y_x $, sending $ \star $ to $ (y_x)_x $. However, they can also be identified with the morphisms $ f: X \to \sum_{x : X} Y_x $ that make the following diagram commute, sending $ x $ to $ y_x $:
\begin{center}
  \begin{tikzcd}
    X \arrow[rd, "\id X"'] \arrow[rr, "f"] & & \sum_{x : X} Y_x \arrow[ld, "\pi_1"]\\
    & X
  \end{tikzcd}
\end{center}
These are morphisms in $ (\SET \downarrow X) $ from $ (X, \id X) $ to $ (\sum_{x : X} Y_x, \pi_1) $. Note that for the terminal morphism $ f: X \to \{ \star \} $, we have $ (X, \id X) = f^*(\{\star\}, \id{\{\star\}}) $. To summarize, we have an equivalence
\[ (\SET \downarrow X)\left(f^*(\{ \star \}, \id{\{\star\}}), (Y_x)_x \right) \simeq (\SET \downarrow {\{\star\}})\left( (\{\star\}, \id{\{ \star \}}), \left( \prod_{x : X} Y_x, ! \right) \right) \]
for $ ! $ the terminal morphism. Now, for an internal indexed family $ f : Y \to X $ and given a family of families $ ((Z_y)_{y : Y_x})_{x : X} $, we can wonder whether we can construct the family of dependent products $ (\prod_{y : Y_x} Z_y)_x $. In $ \SET $, this is definitely possible, and from this, we get an equivalence again
\[ \left(\SET \downarrow \sum_{x : X} Y_x \right)\left(f^*(X, \id X), (Z_y)_{x, y} \right) \simeq (\SET \downarrow X)\left((X, \id X), \left(\prod_{y : Y_x} Z_y\right)_x\right). \]
with $ p : \sum_{x : X} \sum_{y : Y_x} Z_y \to \sum_{x : X} Y_x $ defined as $ p(x, y, z) = (x, y) $. These equivalences suggest an adjunction $ f^* \dashv \prod_{\dots} $. We can use this to define in general
\begin{definition}
  For a category $ C $ and a morphism $ f: Y \to X $, the \iindex{dependent product} along $ f $ is, if it exists, the right adjoint to the pullback functor:
  \begin{center}
    \begin{tikzcd}
      (C \downarrow A) \arrow[r, shift left=2, "f^*"{name=A}] &
      (C \downarrow B) \arrow[l, shift left=2, "\prod_f"{name=B}]
      \ar[from=A, to=B, symbol=\dashv]
    \end{tikzcd}
  \end{center}
\end{definition}
\begin{remark}
  As argued above, we can recover the familiar dependent product $ \prod_{x : X} Y_x $ of a family $ (Y_x)_x $ as the dependent product $ \prod_f (\sum_x Y_x, \pi_1) $ along the terminal morphism $ f: X \to I $. Here we use the equivalence between $ (C \downarrow I) $ and $ C $.
\end{remark}

Now we turn our attention to dependent sums. In $ \SET $, let $ (Y_x)_x $ and $ (Y^\prime_x)_x $ be two families over $ X $ and let $ ((Z_y)_{y : Y_x})_{x : X} $ be a family of families. Let $ f: \sum_{x : X} Y_x \to X $ be the internal representation of $ (Y_x)_x $. A family of maps $ g_x : (\sum_{y : Y_x} Z_y)_x \to Y^\prime_x $ consists of maps $ Z_y \to Y^\prime_x $ for all $ y : Y_x $, so these are maps $ g_y : Z_y \to Y^\prime_{f(y)} $. This gives an equivalence
\[ (\SET \downarrow X)\left(\left( \sum_{y : Y_x} Z_y \right)_x, (Y^\prime_x)_x \right) \simeq \left(\SET \downarrow \left(\sum_{x : X} Y_x \right) \right)((Z_y)_y, f^*((Y^\prime_x)_x)). \]
This, again, suggests an adjunction which we will use as a definition.
\begin{definition}
  For a category $ C $ and a morphism $ f: Y \to X $, the \iindex{dependent sum} along $ f $ is, if it exists, the left adjoint to the pullback functor:
  \begin{center}
    \begin{tikzcd}
      (C \downarrow X) \arrow[r, shift right=2, "f^*"'{name=B}] &
      (C \downarrow Y) \arrow[l, shift right=2, "\sum_f"'{name=A}]
      \ar[from=A, to=B, symbol=\dashv]
    \end{tikzcd}
  \end{center}
\end{definition}

However, note that the conversion from an external to an internal representation in $ \SET $ already contained a dependent sum, which is no coincidence. It turns out that in practice, we will never have a hard time obtaining dependent sums:
\begin{lemma}\label{lem:sum-postcomposition}
  Let $ f : C(Y, X) $ be a morphism in a category. If the pullback functor $ f^*: (C \downarrow X) \to (C \downarrow Y) $ exists, it has a left adjoint given by postcomposition with $ f $.
\end{lemma}
\begin{proof}
  For morphisms $ g: Z \to X $, $ h: W \to Y $, the universal property of the pullback, with the following diagram
  \begin{center}
    \begin{tikzcd}
      W \arrow[r, dashed, "\varphi"'] \arrow[rr, "\psi", bend left] \arrow[dr, "h"'] & f^* Z \arrow[d, "f^* g"'] \arrow[r] \arrow[rd, phantom, "\lrcorner", very near start] & Z \arrow[d, "g"]\\
      & Y \arrow[r, "f"'] & X
    \end{tikzcd}
  \end{center}
  gives an equivalence between morphisms $ \varphi: W \to f^* Z $ that commute with $ h $ and $ f^* g $, and morphisms $ \psi: W \to Z $ that commute with $ h $, $ g $ and $ f $. In other words:
  \[ (C \downarrow X)(h \cdot f, g) \simeq (C \downarrow Y)(h, f^*(g)), \]
  which shows the adjunction.
\end{proof}

Now, let $ g : Y \to X $ be the internal representation of an indexed family $ (Y_x)_x $ and let $ f : X \to I $ be the terminal projection. We have $ \sum_{x : X} Y_x = g \cdot f : Y \to I $. By the equivalence between $ (C \downarrow I) $ and $ C $, we see that the dependent sum of the family $ (Y_x)_x $ is exactly $ Y $. Therefore, our attention is mainly focused on the dependent product.

@@

\begin{definition}
  A \index{cartesian closed!locally}\textit{locally cartesian closed} category is a category $ C $ with pullbacks such that each pullback functor $ f^* $ has a right adjoint.
\end{definition}

Apart from having dependent sums and products, there also is the following theorem that shows the significance of locally cartesian closedness:
\begin{lemma}\label{lem:locally-cartesian-closed}
  A category $ C $ is locally cartesian closed iff $ (C \downarrow X) $ is cartesian closed for each $ X : C $.
\end{lemma}
\begin{proof}
  See the end of Section 1.3 of \autocite{freyd}.
\end{proof}

\begin{remark}\label{rem:pullback-of-projection}
  Note that for $ X, Y, Z : C $ and $ f: C(Y, X) $, the following diagram shows that $ f^* \Delta_X Z \cong \Delta_Y Z $:
  \begin{center}
    \begin{tikzcd}
      Y \times Z \arrow[d, "p_1"] \arrow[r, "f \times \id Z"] & X \times Z \arrow[d, "p_1"] \arrow[r, "p_2"] & Z \arrow[d, "!"]\\
      Y \arrow[r, "f"] & X \arrow [r, "!"] & I
    \end{tikzcd}
  \end{center}
\end{remark}

\begin{lemma}\label{lem:constant-dependent-product}
  For $ Z, X, Y : C $ and $ p_1 : X \times Y \to X $,
  \[ \prod_{p_1} \Delta_{X \times Y} Z \cong \Delta_X Z^Y \]
\end{lemma}
\begin{proof}
  First of all, note that $ (C \downarrow X \times Y) \cong ((C \downarrow X) \downarrow \Delta_X Y) $. Also note that the composite morphism $ (X \times Y) \times Z \xrightarrow{p_1} X \times Y \xrightarrow{p_1} X $ is the element $ \Delta_X (Z \times Y) : (C \downarrow X) $.

  By Proposition 1.34 in \autocite{freyd}, $ \prod_{p_1} \Delta_{X \times Y} Z $ is given as the following pullback:
  \begin{center}
    \begin{tikzcd}
      \prod_{p_1} \Delta_{X \times Y} Z \arrow[r] \arrow[d] & (\Delta_X Y \times Z)^{\Delta_X Y} \arrow[d, "(\Delta_X p_1)^{\Delta_X Y}"]\\
      X \arrow[r] & (\Delta_X Y)^{\Delta_X Y}
    \end{tikzcd}
  \end{center}

  By Lemma \ref{lem:delta-exponentials}, $ \Delta_X $ preserves exponential objects, so the morphism on the right is $ \Delta_X p_1^Y : (C \downarrow X)(\Delta_X (Y \times Z)^Y, \Delta_X Y^Y) $. However, we have an isomorphism $ (Y \times Z)^Y \cong Y^Y \times Y^Z $, and then the morphism on the right becomes
  \[ \Delta_X p_1 : (C \downarrow X)(\Delta_X(Y^Y \times Y^Z), \Delta_X Y^Y) \]
  We also have an isomorphism $ X \cong \Delta_X I $. Then by Lemma \ref{lem:delta-limits} and Remark \ref{rem:pullback-of-projection}, the pullback of this diagram is $ \Delta_X Z^Y $:
  \begin{center}
    \begin{tikzcd}
      \Delta_X Z^Y \arrow[r] \arrow[d] & \Delta_X (Y^Y \times Z^Y) \arrow[d, "\Delta_X p_1"]\\
      \Delta_X I \arrow[r] & \Delta_X Y^Y
    \end{tikzcd}
  \end{center}
\end{proof}

\section{(Weakly) terminal objects}
\begin{definition}
  If a category has an object $ I $, such that there is a (not necessarily unique) morphism to it from every other object in the category, $ I $ is said to be a \iindex{weakly terminal object}.
\end{definition}

\begin{definition}
  Let $ C $ be a category with terminal object $ I $. For an object $ X: C $, a \iindex{global element} of $ X $ is a morphism $ f: C(I, X) $.
\end{definition}
